---
title: "MA8701 Advanced methods in statistical inference and learning"
author: "Mette Langaas IMF/NTNU"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    toc: yes
    toc_depth: 3
    latex_engine: xelatex
  ioslides_presentation: default
  beamer_presentation:
    slide_level: 1
    keep_tex: yes
    latex_engine: xelatex
  html_document:
    toc: yes
    toc_float: yes
    code_download: yes
    toc_depth: 3
subtitle: 'L7: Random forest, Super Learner, Hyperparameter tuning'
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE,echo=FALSE}
suppressPackageStartupMessages(library(knitr))
knitr::opts_chunk$set(echo = FALSE, message=FALSE,warning = FALSE, error = FALSE)
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(rpart))
suppressPackageStartupMessages(library(rpart.plot))
suppressPackageStartupMessages(library(randomForest))
suppressPackageStartupMessages(library(DAAG))
suppressPackageStartupMessages(library(tree))
suppressPackageStartupMessages(library(MASS))
suppressPackageStartupMessages(library(GGally))
suppressPackageStartupMessages(library(corrplot))
suppressPackageStartupMessages(library(caret)) #for confusion matrices
suppressPackageStartupMessages(library(pROC)) #for ROC curves
suppressPackageStartupMessages(library(SuperLearner))
suppressPackageStartupMessages(library(tidymodels))
#suppressPackageStartupMessages(library(SuperLearner))
```

# Ensembles - third act

## So far

**L5**:

-   Trees
-   Many trees with bootstrap aggregation
-   Many trees into a random forest (not finished)

**L6**: with Berent Å. S. Lunde (UiB)

-   Boosting
-   Xgboost
-   Avoiding hyperparameter tuning in xgboost

------------------------------------------------------------------------

## Outline

-   Ensembles
-   Finishing the random forest from
    [L5](http://htmlpreview.github.com/?https://github.com/mettelang/MA8701V2021/blob/main/Part2/L5.html)
-   Super Learner
-   Hyperparameter tuning

Left for L8: How to use statistical inference to give CI and compare
predictions from different methods.

------------------------------------------------------------------------

# Ensembles - overview

(ELS Ch 16.1)

With ensembles we want to build *one prediction model* which combines
the strength of *a collection of models*.

These models may be simple base models - or more elaborate models.

We have studied bagging - where we take a simple average of the
prediction from many models (or majority vote), and the base models can
be trees - or other type of models.

Random forest is a version of bagging, with trees made to be different
(decorrelated).

------------------------------------------------------------------------

We have studied boosting, where the models are trained on sequentially
different data - from residuals or gradients of loss functions - and the
ensemble members cast weighted votes. We have in particular looked into
the xgboost variant of boosting, and also evaluated possible parameters
to tune to optimize performance.

In L7 we also look into an ensemble built of elaborate models with the
Super Learner and study some possible methods for tuning
hyperparameters.

# Random forest

(ELS Ch 15) see learning material in L5.

# Super Learner

(References: Le Dell (2015) Section 2.2 and van der Laan et al (2016),
and Polley et al (2011))

Why do we want to study the Super Learner?

The Super Learner or *generalized stacking* is an algorithm that
combines

-   multiple, (typically) diverse prediction methods (learning
    algorithms) called *base learners* into a
-   a *metalearner* - which can be seen as a *single* method.

------------------------------------------------------------------------

## Development:

-   1992: stacking introduce for neural nets by Wolpert
-   1996: adapted to regression problems by Breiman - but only for one
    type of methods at once (cart with different number of terminal
    nodes, glms with subset selection, ridge regression with different
    ridge penalty parameters)
-   2006: proven to have asymptotic theoretical oracle property by van
    der Laan, Polley and Hubbard.

------------------------------------------------------------------------

## Ingredients:

-   *Training data* (level-zero data) $O_i=(X_i,Y_i)$ of $N$ i.i.d
    observations.
    
-   A total of $L$ *base learning algorithms* $\Psi^l$ for
    $l=1,\ldots,L$, each from some algorithmic class and each with a
    specific set of model parameters.
    
-   A *metalearner* ${\bf \Phi}$ is used to find an *optimal
    combination* of the $L$ base learners.

------------------------------------------------------------------------

## Algorithm

**Step 1: Produce level-one data** ${\bf Z}$

a)  Divide the training data ${\bf X}$ randomly into $V$ roughly-equally
    sized validation folds ${\bf X}_{(1)},\ldots,{\bf X}_{(V)}$. $V$ is
    often 5 or 10. (The responses ${\bf Y}$ are also needed.)

b)  For each base learner $\Psi^l$ perform $V$-fold cross-validation to
    produce prediction.

This gives the level-one data set ${\bf Z}$ consisting prediction of all
the level-zero data - that is a matrix with $N$ rows and $L$ columns.

**What could the base learners be?**

------------------------------------------------------------------------

"Any" method that produces a prediction - "all" types of problems.

-   linear regression
-   lasso
-   cart
-   random forest with mtry=value 1
-   random forest with mtry=value 2
-   xgboost with hyperparameter set 1
-   xgboost with hyperparameter set 2
-   neural net with hyperparameter set 1

------------------------------------------------------------------------

**Step 2: Fit the metalearner**

a)  The starting point is the level-one prediction data ${\bf Z}$
    together with the responses $(Y_1,\ldots ,Y_N)$.
b)  The metalearner is used to estimate the weights given to each base
    learner: $\hat{\eta_i}=\alpha_1 z_{1i}+ \cdots + \alpha_L z_{Li}$.

**What could the metalearner be?**

------------------------------------------------------------------------

* the mean (bagging)
* constructed by minimizing the 
  + squared loss (ordinary least squares) or
  +  non-negative least squares
* ridge or lasso regression
* constructed by minimizing 1-ROC-AUC

------------------------------------------------------------------------

(Class notes: Study Figure 3.2 from Polley et al)

------------------------------------------------------------------------

**Step 3: Re-estimate base learners and combine into superlearner on
full training data**

a)  Fit each of the $L$ base learners to the full training set.
b)  The *ensemble fit* consists the $L$ base learner fits together with
    the metalearner fit.

**Step 4: Using the ensemble for prediction**

For a new observaton ${\bf x}^*$

a)  Use each of the $L$ base learners to produce a prediction
    ${\bf z}^*$, and
b)  feed this to the metalearner-fit to produce the final prediction
    $y^*$.

------------------------------------------------------------------------

## The metalearning

Some observations

-   The term *discrete super learner* is used if the base learner with
    the lowest risk (i.e. CV-error) is selected.
-   Since the predictions from multiple base learners may be highly
    correlated - the chosen method should perform well in that case
    (i.e. ridge and lasso).
-   When minimizing the squared loss it has been found that adding a
    non-negativity constraint $\alpha_l\le 0$ works well,
-   and also the additivity constraint $\sum_{l=1}^L \alpha_l=1$ - the
    ensemble is a *convex combination* of the base learners.
-   Non-linear optimization methods may be employed for the metalearner
    if no existing algorithm is available
-   Historically a regularized linear model has "mostly" been used
-   For classification the logistic response function can be used on the
    linear combination of base learners (Figure 3.2 Polley).

------------------------------------------------------------------------

## Examples

### Simulations examples

(Class notes: Study Figure 3.3 and Table 3.2 from Polley et al)

### Real data

(Class notes: Study Figure 3.4 and Table 3.3 from Polley et al. RE=MSE
relative to the linear model OLS.)

------------------------------------------------------------------------

## Theoretical result
(Le Dell 2015, page 6)

-   Oracle selector: the estimator among all possible weighted
    combinations of the base prediction function that minimizes the risk
    under the *true data generating distribution*.
    
    
-   The *oracle result* was established for the Super Learner by van der
    Laan et al (2006).
    
    
-   If the *true prediction function* cannot be represented by a
    combination of the base learners (available), then "optimal" will be
    the closest linear combination that would be optimal if the true
    data-generating function was known.
    
    
-   The oracle result require an *uniformly bounded loss function*.
    Using the convex restriction (sum alphas =1) implies that if each
    based learner is bounded so is the convex combination. In practice:
    truncation of the predicted values to the range of the outcome in
    the training set is sufficient to allow for unbounded loss functions


------------------------------------------------------------------------

## Uncertainty in the ensemble

(Class notes: Study "Road map" 2 from Polley et al)

-   Add an outer (external) cross validation loop (where the super
    learner loop is inside). Suggestion: use 20-fold, especially when
    small sample size.
    
-   Overfitting? Check if the super learner does as well or better than
    any of the base learners in the ensemble.
    
-   Results using *influence functions* for estimation of the variance
    for the Super Learner are based on asymptotic variances in the use
    of $V$-fold cross-validation (see Ch 5.3 of Le Dell, 2015)

------------------------------------------------------------------------

## Other issues

-   Many different implementations available, and much work on parallell
    processing and speed and memory efficient execution.
    
-   Super Learner implicitly can handle hyperparameter tuning by
    including the same base learner with different model parameter sets
    in the ensemble.
    
-   Speed and memory improvements for large data sets involves
    subsampling, and the R `subsemble` package is one solution, the H2O
    Ensemble project another.

------------------------------------------------------------------------

## R example

Code is copied from [Guide to SuperLearner](https://cran.r-project.org/web/packages/SuperLearner/vignettes/Guide-to-SuperLearner.html) and the presentation
follows this guide. The data used is the Boston housing dataset from
`MASS`, but with the median value of a house dichotomized into a
classification problem.

Observe that only 150 of the 560 observations is used (to speed up
things, but of cause that gives less accurate results).

```{r,echo=TRUE}
data(Boston, package = "MASS")
#colSums(is.na(Boston)) # no missing values
outcome = Boston$medv
# Create a dataframe to contain our explanatory variables.
data = subset(Boston, select = -medv)
#Set a seed for reproducibility in this random sampling.
set.seed(1)
# Reduce to a dataset of 150 observations to speed up model fitting.
train_obs = sample(nrow(data), 150)
# X is our training sample.
x_train = data[train_obs, ]
# Create a holdout set for evaluating model performance.
# Note: cross-validation is even better than a single holdout sample.
x_holdout = data[-train_obs, ]
# Create a binary outcome variable: towns in which median home value is > 22,000.
outcome_bin = as.numeric(outcome > 22)
y_train = outcome_bin[train_obs]
y_holdout = outcome_bin[-train_obs]
table(y_train, useNA = "ifany")
```

Then checking out the possible functions and how they differ from their
"original versions".

```{r,echo=TRUE}
listWrappers()
# how does SL.glm differ from glm? obsWeight added to easy use the traning fold in the CV and returns a prediction for new observarions
SL.glm
# min and not 1sd used, again obsWeights, make sure model matrix correctly specified
SL.glmnet
```

The fitting lasso to check what is being done. The default metalearner
is "method.NNLS" (both for regression and two-class classification -
probably then for linear predictor NNLS?).

```{r,echo=TRUE}
set.seed(1)
sl_lasso=SuperLearner(Y=y_train, X=x_train,family=binomial(),SL.library="SL.glmnet")
sl_lasso
#str(sl_lasso)
sl_lasso$cvRisk
```

Now use lasso and randomforest, and also add the average of ys just as
the benchmark.

```{r,echo=TRUE}
set.seed(1)
sl=SuperLearner(Y=y_train, X=x_train,family=binomial(),SL.library=c("SL.mean","SL.glmnet","SL.randomForest"))
sl
sl$times$everything
```

Our ensemble give weight 0.13 to lasso and 0.86 to the random forest.
(The guide used a different implementation of the random forest called
ranger, and got 0.02 and 0.98.)

Predict on the part of the dataset not used for the training.

```{r,echo=TRUE}
pred=predict(sl,x_holdout=x_holdout,onlySL=TRUE)
str(pred)
summary(pred$pred)
summary(pred$library.predict)
```

Add now an external cross-validation loop - only using the training
data. Here the default $V=10$ is used for the inner loop, and we set the
value for the outer loop (here $V=3$ for speed).

```{r,echo=TRUE}
system.time({cv_sl=CV.SuperLearner(Y=y_train, X=x_train,V=10,family=binomial(),SL.library=c("SL.mean","SL.glmnet","SL.randomForest"))})
summary(cv_sl)
```

See the guide for more information on running multiple versions of one
base learner, and parallellisation.

# Choosing hyperparameters

-   What are *hyperparameters*?
-   Which hyperparameters have we encountered in the course so far?

------------------------------------------------------------------------

*Hyperparameters* are parameters than cannot be directly estimated from
data. This may be model parameters (the $\lambda$ in lasso) or
parameters that influence the fitting of the model (e.g. related to some
optimization algorithm).

We have already studied hyperparameter tuning for the lasso, ridge,
elastic net, random forest, and boosting - with the use of
cross-validation of some loss function for a predefined set of
hyperparameter values.

The use of ensembles like the Super Learner may be seen as an
alternative to hyperparameter tuning.

------------------------------------------------------------------------

Overview of advices from Berent Å.S. Lunde on tuning parameters in
xbgoost (written out after the lecture and approved by Berent):

**Ways to speed-up computation:**

-   Higher learning-rate, then tune the number of boosting iterations.
-   When tuning, do not use too high k in k-fold CV (for both speed and
    also to avoid high variance), or drop CV and use a validation set.
-   Speedups with histogram algorithms of order $n$ (avoid exact
    enumeration of all splits, nlogn).
-   Use sparsity when possible!

------------------------------------------------------------------------

**Comments on hyperparameters:**

-   Learning rate (eta in xgb): Set as low as computation times allow.
    Typically in between $0.01$ and $0.1$ is sufficient.
-   Number of trees (boosting iterations): Very much affected by the
    learning rate. Most important parameter to tune, given a learning
    rate. Maximum depth of trees: start low, then increase. High
    values takes significantly longer to train. Think about the problem,
    and the number of possible interaction effects. If max-depth = J,
    then interactions among J-1 features is possible. How much is
    needed?
-   Gamma: Very traditional tree-hyperparameter to tune. Tune.
-   Colsampling for tree is usually sufficient.
-   subsample: in 0.5-0.9
-   Min child weight: Something I usually do not think about. Default=1
    (low) works. For me

Every problem is different, and there will exist exceptions to the above
guidelines. *I look forward to a world without hyperparameters.* (Berent
Å.S. Lunde)

------------------------------------------------------------------------

The hyperparameters may be continuous (penalty parameter), discrete
(number of layers in a neural network, applying early stopping or not)
or categorical (choose between different optimizers).

The choice of hyperparameters is important, and will often directly
affect the model complexity, and unwise choices may lead to overfitting.

------------------------------------------------------------------------

Hyperparametertuning is performed using a separate validation set or by
cross-validation. Different loss functions or selection criteria may be
used (MSE, AUC, misclassification rate, ...).

The hyperparameter tuning is often referred to as a black-box
optimization because we (usually) only calculate loss function values
(with CV) and do not get to compute gradients.

What may be challenges with hyperparameter optimization?

------------------------------------------------------------------------

Some challenges with hyperparameter optimization (Feurer and Hutter,
Ch1):

-   expensive evaluation of the model under study (large networks, large
    data sets)
-   unclear which of possibly many hyperparameters that need to be
    selected carefully (refer to the discussion for xgboost)
-   gradient of selection criterion with respect to the hyperparameters
    not (generally) available, and criterion not convex or smooth in the
    hyperparameters
-   and the need for external validation or CV

------------------------------------------------------------------------

There exist many ways to *group* methods for hyperparameter tuning. One
way to look at this is (Kuhn and Silge, 2021, Ch 12)

-   grid search: specify a set of possible values a priori and
    investigate only these values, choose the value where the chosen
    selection criterion is optimal. This is also called "model free
    methods".
-   iterative search: start with a set of values, fit/evaluate some
    (surrogate) model (might also be the loss function), and based on
    this choose new values to evaluate next.

For grid search also methods for *speeding up calculations* exists - for
example by stopping evaluation at a grid point where the loss is seen to
be high after some CV-folds, for example the method of *racing*
described by Kuhn and Silge, Ch 13.4.

------------------------------------------------------------------------

(Class notes: see example from Kuhn and Silge, 2021, Spacefilling grid
vs global search)

------------------------------------------------------------------------

## Surrogate methods

We will look at two types of surrogate models: Bayesian regression with
Gaussian processes (in Bayesian optimization) and regression-type models
in response surface methods (presented by group 2).

## Design of experiments and response surface methodology

Article presentation by group 2.

G. A. Lujan-Moreno, P. R. Howard, O. G. Rojas and D. C. Montgomery
(2018): Design of experiments and response surface methodology to tune
machine learning hyperparameters, with a random forest case- study.
Expert Systems with Applications. 109, 195-205.

---

## Bayesian optimization

Bayesian optimization is an iterative method - where we start with
evaluating some loss function at some predefined set of points in the
hyperparameter space. New position in the hyperparameter space are
chosen iteratively.

Two key ingredients:

-   a surrogate model (we will only look at Bayesian regression with
    Gaussian processes) to fit to the observed values of the loss
    function in the hyperparameter space
-   an *acquisition* function to decide a new point in the
    hyperparameter space to evaluate next

------------------------------------------------------------------------

Underlying idea: given some "observations" in the hyperparameter space,
the task is to decide where to place a new point. We should try a point
where:

-   we expect a good value and/or
-   we have little information so far

To do that we need information on both expected value *and* variance -
or preferably the distribution of the loss function for our problem.

We now look at the multivariate Gaussian distribution and conditional
distribution, a Gaussian process

------------------------------------------------------------------------


### Gaussian processes

(Eidsvik 2017, page 6-7, note in TMA4265)

A Gaussian process is defined for

-   times or locations $x_i$, $i=1,\ldots,n$ in $\Re^d$, where
-   $Y_i=Y(x_i)$ is a random variable at $x_i$
-   such that ${\bf Y}=(Y_1,\ldots,Y_n)$ is multivariate Gaussian.

The process is *first order (mean) stationary* if $\text{E}(Y(x))=\mu$
for all $x$, and this can be extended to depend on covariates.

The process is *second order stationary* if $\text{Var}(Y(x))=\sigma^2$
for all $x$ and the correlation $\text{Corr}(Y(x),Y(x'))$ only depends
on differences between $x$ and $x'$.

The multivariate Gaussian distribution is defined by the mean and
covariance alone.

------------------------------------------------------------------------

#### Correlation functions

(Eidsvik 2017, page 7, Frazier 2018, Ch 3.1)

Correlation functions are also referred to as *kernels*.

We assume that points at positions close to each other have a stronger
correlation than point far apart.

**Power exponential or Gaussian kernel**
<!-- $$\text{Corr}(Y(x),Y(x'))=\alpha_0 \exp(-\sum_{j=1}^d \alpha_j (x_j-x_j')^2) -->
<!-- $$ -->
$$ \text{Corr}(Y(x),Y(x'))=\exp(-\phi_G \Vert x-x' \Vert ^2)$$ where the
L2 distance is used and $\phi_G$ is a parameter that determine the decay
in the correlations.

<!-- (Class notes: study Figure 2 of Frazier 2018 for plots for Gaussian kernel.) -->

------------------------------------------------------------------------

**Matern-type kernel**

<!-- $$\text{Corr}(Y(x),Y(x´))=\alpha_0 \frac{2^{1-\nu}}{\Gamma(\nu)} (\sqrt{2\nu}\sum_{j=1}^d \alpha_j (x_j-x_j'))^{\nu} K_{\nu}(\sum_{j=1}^d \alpha_j (x_j-x_j'))  -->

<!-- $$with the modified Bessel function $K_{\nu}$ and additional  -->

$$\text{Corr}(Y(x),Y(x´))=(1+\phi_M \Vert x - x' \Vert)\exp(-\phi_M \Vert x - x' \Vert)$$

now with decay-describing parameter $\phi_M$.

The parameters of the kernels need to be estimated, see Ch 3.2 of
Frazier 2018 (who use a slightly different parameterization). We will
just assume that these parameters are known.

(Class notes: study Figure 4 of Eidsvik, 2018.)

------------------------------------------------------------------------

#### From correlations into covariance matrix

For simplicty assume that $d=1$. The number of positions to consider is
$n$.

To get from correlation function to a $n \times n$ covariance matrix
first construct a $n \times n$ matrix of distances for each pair of
positions, denote this ${\bf H}$.

For the Matern-type correlation function the covariance matrix can then
be written

$$ \Sigma=\sigma^2 (1+\phi_M {\bf H}) \otimes \exp(-\phi_M {\bf H}))$$
where $\otimes$ is elementwise multiplication.

See Eidsvik (2018, Ch 3.2 and 3.3) for how to build covariance matrices
in an efficient way.

------------------------------------------------------------------------

### Multivariate normal distribution

aka multivariate Gaussian distribution. Known from TMA4265 and TMA4267.

The random vector $\mathbf{Y}_{p\times 1}$ is multivariate normal $N_p$
with mean $\mathbf{\mu}$ and (positive definite) covariate matrix
$\Sigma$. The pdf is:

$$f(\mathbf{Y})=\frac{1}{(2\pi)^\frac{p}{2}|\Sigma|^\frac{1}{2}} \exp\{-\frac{1}{2}(\mathbf{Y}-\mathbf{\mu})^T\Sigma^{-1}(\mathbf{Y}-\mathbf{\mu})\}$$

Six useful properties of the mvN - we need number 6.

Let $\mathbf{Y}_{(p\times 1)}$ be a random vector from
$N_p(\mathbf{\mu},\Sigma)$.

1.  The grapical contours of the mvN are ellipsoids (can be shown using
    spectral decomposition).
2.  Linear combinations of components of $\mathbf{Y}$ are (multivariate)
    normal (can be easily proven using moment generating functions MGF).
3.  All subsets of the components of $\mathbf{Y}$ are (multivariate)
    normal (special case of the above).

------------------------------------------------------------------------

4.  Zero covariance implies that the corresponding components are
    independently distributed (can be proven using MGF).
5.  $\mathbf{A}\Sigma\mathbf{B}^T=\mathbf{0} \Leftrightarrow \mathbf{A}\mathbf{Y}$
    and $\mathbf{B}\mathbf{Y}$ are independent.
6.  The conditional distributions of the components are (multivariate)
    normal.
    $$\mathbf{Y}_2 \mid (\mathbf{Y}_1=\mathbf{Y}_1) \sim N_{p2}(\mathbf{\mu}_2+\Sigma_{21}\Sigma_{11}^{-1} (\mathbf{Y}_1-\mathbf{\mu}_1),\Sigma_{22}-\Sigma_{21}\Sigma_{11}^{-1}\Sigma_{12}).$$

------------------------------------------------------------------------

### Acquisition function: Expected improvement

(Frazier 2018 page 7)

Thought experiment:

1)  we have evaluated our function at all possible points $x$, and must
    return a solution based on what we already have evaluated. If the
    evaluation is noise-less we need to return the point with the
    largest observed value $f$.

2)  Correction: We may perform one more evaluation. If we choose $x$ we
    observe $f(x)$, and the best point before that was $f^{*}_n$. The
    improvement at the new observation is then $$ \max(f(x)-f^{*}_n,0)$$

------------------------------------------------------------------------

3)  We define the *expected improvement* as

$$ \text{EI}_n(x)=\text{E}_n[\max(f(x)-f^{*}_n,0)]$$

where the expectation is taken at the posterior distribution given that
we have evaluated $f$ at $n$ observations $x_1,\ldots, x_n$, and the
posterior distribution is that $f$ conditional on
$x_1,\ldots,x_n,y_1,\ldots,y_n$ is normal with mean $\mu_n(x)$ and
variance $\sigma^2_n(x)$.

------------------------------------------------------------------------

4)  How to evaluate the expected improvement? Integration by parts gives

$$ \text{EI}_n(x)=\max(\mu_n(x)-f^{*}_n,0)])+\sigma_n(x) \phi(\frac{\mu_n(x)-f^{*}_n}{\sigma_n(x)}) $$
$$-\text{abs}(\mu_n(x)-f^{*}_n) \Phi(\frac{\mu_n(x)-f^{*}_n}{\sigma_n(x)})$$

$\mu_n(x)-f^{*}_n$ is expected proposed vs previously best

5)  We choose to evaluate the point with the largest expected
    improvement

$$ x_{n+1}=\text{argmax}\text{EI}_n(x)$$

Is often found using quasi-Newton optimization.

------------------------------------------------------------------------

### Algorithm for Bayesian optimization of a function $f$

(Frazier 2018, page 3, noise-free evaluation)

------------------------------------------------------------------------

Place a Gaussian process prior on $f$.

Observe $f$ at $n_0$ points from some experimental design. Set $n=n_0$.

**while** $n \le N$ do

Update the posterior on f with all available data

Let $x_n$ be a maximizer of the acquisition function over $x$, computed
using the current posterior

Observe $y_n=f(x_n)$

Increment $n$

**end while**

Return a solution: a point with largest $f(x)$ or the point with the
largest posterior mean

------------------------------------------------------------------------

What does the steps mean?

-   Gaussian prior: choose (estimate?) mean and correlation function for
    the problem.
-   Observe $n_0$ points: calculate the loss function at each of the
    points (remark: we have noise)
-   Update the posterior: calculate the conditional distribution for $f$
    for a new point given the observed loss at all previously observed
    points
-   Acquisition function: find $\text{argmax}\text{EI}_n(x)$.

(Class notes: Figure 1 of Frazier 2018.)

------------------------------------------------------------------------

-   For a point $x$ we model the distribution of $f(x)$,
-   which is normally distributed with mean $\mu_n(x)$ and variance
    $\sigma_n^2(x)$. The mean and variance is found from the conditional
    distribution.
-   With 95% credibility interval $\mu_n(x)\pm 1.95 \sigma_n(x)$.
-   The width of the credibility interval at observations is 0.

------------------------------------------------------------------------

## Extension

What is the objection function is not observed noise-less?

Independent normal error term $\varepsilon$ can be added to the previously defined $Y=f(x)$ to make a new $Y=f(x)+\varepsilon$.
This (only) adds a diagonal term to the covariance matrix, and it is common to assume that the variance is the same for all $x$ and treat the variance as a hyperparameter.

---

### Example

(Kuhn and Silge, Ch 14, the example is for SVM)

First just grid search to test what is best value for `mtry`

```{r,echo=TRUE}
data(Boston, package = "MASS")
# first using a grid
tune_grid <- expand.grid(
  mtry = (1:13))
#  ntree=seq(100,500,length=10)) # how to also include ntree? primary only mtry, how to define secondary?
tune_control <- caret::trainControl(
  method = "oob", # cross-validation #eller cv
  #number = 3, # with n folds 
  verboseIter = FALSE, # no training log
  allowParallel = FALSE # FALSE for reproducible results 
)
rf_tune <- caret::train(
  medv~crim+zn+indus+chas+nox+rm+age+dis+rad+tax+ptratio+black+lstat, 
  data=Boston,
  na.action=na.roughfix,
  trControl = tune_control,
  tuneGrid = tune_grid,
  method = "rf", # rf is randomForest, checked at #vhttp://topepo.github.io/caret/train-models-by-tag.html#Random_Forest
  verbose = TRUE
)
tuneplot <- function(x, probs = .90) {
  ggplot(x) +
    coord_cartesian(ylim = c(quantile(x$results$RMSE, probs = probs), min(x$results$RMSE))) +
    theme_bw()
}
tuneplot(rf_tune)
rf_tune$bestTune
```

------------------------------------------------------------------------

The R the function `tune_bayes` is available in the package `tune`, and
requires that the analyses is done with a workflow. Default in the GP is
exponential correlation function, but first we try the Matern.

```{r,echo=TRUE,eval=FALSE}
tree_rec <- recipe(medv~crim+zn+indus+chas+nox+rm+age+dis+rad+tax+ptratio+black+lstat, data = Boston)

tune_spec <- rand_forest( # parsnip interface to random forests models
  mode="regression",
  mtry = tune(),
  trees = tune(),
#  min_n = tune()
) %>%
#  set_mode("regression") %>%
#  set_engine("ranger",objective="reg:rmse") # errors with ranger
  set_engine("randomForest") # randomforest ok

tune_wf <- workflow() %>%
  add_recipe(tree_rec) %>%
  add_model(tune_spec)

tune_param <- tune_spec%>% 
  parameters%>% 
  update(mtry=mtry(c(1L,13L)),trees=trees(c(100L,500L)))

vfold  <- vfold_cv(Boston, v = 5)
# then trying BO
ctrl <- control_bayes(verbose = TRUE)
bayesres<- tune_bayes(tune_wf,
    resamples = vfold,
    #metrics = rmse,
    corr=list(type="matern",nu=5/2), 
    #default in corr_mat(GPfit) is "exponential" power 1.95
    initial = 10,
    param_info = tune_param,
    iter = 10,
    objective=exp_improve(),
    control = ctrl
  )
dput(bayesres,"bayesres.dd")
```

```{r}
bayesres=dget("bayesres.dd")
show_best(bayesres,n=10)
autoplot(bayesres,type="performance")
autoplot(bayesres,type="parameters")
#names(bayesres)
```

------------------------------------------------------------------------

Here we try the default exponential correlation.

```{r,echo=TRUE, eval=FALSE}
bayesres2<- tune_bayes(tune_wf,
    resamples = vfold,
    #metrics = rmse,
    #corr=list(type="matern",nu=5/2), 
    #default in corr_mat(GPfit) is "exponential" power 1.95
    initial = 10,
    param_info = tune_param,
    iter = 10,
    objective=exp_improve(),
    control = ctrl
  )
dput(bayesres2,"bayesres2.dd")
```

```{r,echo=TRUE}
bayesres2=dget("bayesres2.dd")
show_best(bayesres2,n=10)
autoplot(bayesres2,type="performance")
autoplot(bayesres2,type="parameters")
```

------------------------------------------------------------------------

### Suggested software

(Frazier 2018, Ch 6)

-   R: DiceOptim (on CRAN)

-   R: tune_bayes in `tune` (also CRAN)

-   Python: Spearmint <https://github.com/HIPS/Spearmint>

-   Python: GPyOpt <https://github.com/SheffieldML/GPyOpt>

-   Python: GPFlow (Tensorflow) <https://github.com/GPflow/GPflow> and
    GPYTorch (PyTorch) <https://github.com/cornellius-gp/gpytorch>

------------------------------------------------------------------------


# References

## Super Learner

-   Erin Le Dell (2015): [Scalable Ensemble Learning and Computationally
    Efficient Variance Estimation. PhD Thesis, University of California,
    Berkeley.](https://escholarship.org/uc/item/3kb142r2)

-   Mark J. van der Laan, Eric C. Polley, Alan E. Hubbard (2007): Super
    Learner, Statistical Applications in Genetics and Molecular Biology,
    6, 1, 25

-   Eric C. Polley, Sherri Rose, Mark J. van der Laan (2011): Super
    Learning. Chapter 3 of M. J. van der Laan and S. Rose. Targeted
    Learning, Springer.

## Ensembles

-   [ELS] The Elements of Statistical Learning: Data Mining, Inference,
    and Prediction, Second Edition (Springer Series in Statistics, 2009)
    by Trevor Hastie, Robert Tibshirani, and Jerome Friedman.
    [Ebook](https://web.stanford.edu/~hastie/Papers/ESLII.pdf).

## Hyperparameter tuning

-   M. Feurer and F. Hutter (2019). In F. Hutter et al (eds.) Automated
    Machine Learning. The Springer Series on Challenges in Machine
    Learning.

-   Jo Eidsvik (2017): Introduction to Gaussian processes, note to
    TMA4265.

-   Peter I. Frazier (2018): A tutorial on Bayesian optimization.
    arxiv <https://arxiv.org/abs/1807.02811>

-   Max Kuhn and Julia Silge Version 0.0.1.9008 (2021-02-15) Tidy
    modelling with R. <https://www.tmwr.org/>

-   Roger Gosse, University of Toronto: CSC321 Lecture 21: Bayesian
    Hyperparameter Optimization.
    <http://www.cs.toronto.edu/~rgrosse/courses/csc321_2017/slides/lec21.pdf>

-   Max Kuhn (2020). caret: Classification and Regression Training. R
    package version 6.0-86. <https://CRAN.R-project.org/package=caret>
